---
title: 'INFO201 Problem Set: Rmarkdown and plotting'
author: "Anika Razdan"
date: "2/17/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
#Call libraries
library(ggplot2)
library(data.table)
library(dplyr)

#Set working directory 
setwd("/Users/stlp/Documents/INFO201/")

#Turn off scientific notation 
options(scipen=999)
```

# 1 Load and Check data 

**1. A new markdown file has been created with an appropriate title.**

**2. Load data.**
```{r echo=FALSE, results='hide'}
#Read in data 
dir <- "/Users/stlp/Documents/INFO201/INFO-201/"
df <- read.csv(paste0(dir,"gapminderdata.csv"))
dim(df)
```

We have 13,055 rows and 25 columns. 


**3. Print a small sample of the data.**
```{r echo=FALSE}
head(df,5)
```


# 2 Descriptive statistics 

**1. How many countries are there in the dataset?**
```{r echo=FALSE, results='hide'}
length(unique(df$name))
length(unique(df$iso2))
length(unique(df$iso3))
```

There are 250 countries for *names*, 249 countries for *iso2*, and 253 countries for *iso3*. 


**2. Why are there more names than iso-2 codes and even more iso3-codes?**
a. 




**3. Minimum and maximum year in the data**
```{r echo=FALSE, results='hide'}
min(df$time, na.rm = TRUE)
max(df$time, na.rm = TRUE)
```

The minimum year is 1960 and the maximum year is 2019. 



# 3 CO2 emissions

**1. Analyze CO2 data**
```{r echo=FALSE}
df %>% group_by(time) %>% 
  summarise(co2_na = sum(is.na(co2)),
            co2_PC_na = sum(is.na(co2_PC))) %>% 
  print(n = 1e3)
```

The years 2017, 2018, and 2019 have the most missing data for *co2* and *co2_PC*. 



**2. Plot total CO2 emissions over time for US, China, India, and countries of choice**

```{r echo=FALSE}
df %>% filter(name==c("United States of America", "China", "India", "Uruguay", "France")) %>%
       ggplot(aes(x=time, y=co2, color=name)) + geom_point() + geom_line() +
          xlab("Year") + ylab("CO2 Emissions") + guides(color=guide_legend(title="Countries"))
```



**3. Plot total CO2 emissions per capita over time for US, China, India, and countries of choice**

```{r echo=FALSE}
df %>% filter(name==c("United States of America", "China", "India", "Uruguay", "France")) %>%
       ggplot(aes(x=time, y=co2_PC, color=name)) + geom_point() + geom_line() +
          xlab("Year") + ylab("CO2 Emissions Per Capita") + guides(color=guide_legend(title="Countries"))
```



**4. Average CO2 emissions per capita across the continents **

```{r echo=FALSE}
df %>% group_by(region, time) %>% filter(region!="") %>%
  summarise(co2_PC_avg = mean(co2_PC, na.rm=TRUE)) %>% 
  print(n = 1e3)
```


**5. Barplot for average CO2 emissions per capita across continents in 1960 and 2016**

```{r echo=FALSE}
df %>% filter(time==c(1960, 2016) & region!="") %>% 
  group_by(region, time) %>% 
  summarise(co2_PC_avg = mean(co2_PC, na.rm=TRUE)) %>%
  ggplot(aes(x=region, y=co2_PC_avg, fill=factor(time))) + geom_bar(position="dodge", stat="identity") +
  xlab("Continent") + ylab("Average CO2 per capita") + guides(fill=guide_legend(title="Year"))

```


**6. Answer this **






# 4 GDP per capita

**1. Scatterplot of GDP per capita vs. life expectancy by country in 1960**
```{r echo=FALSE}
df %>% filter(time==1960 & region!="") %>%
  ggplot(aes(x=GDP_PC, y=lifeExpectancy,color=region, size=totalPopulation)) + geom_point() + 
  xlab("GDP per capita") + ylab("Life Expectancy") + scale_size(guide="none") +
  guides(color=guide_legend(title="Continent")) +
  ggtitle("GDP per capita vs. life expectancy by country in 1960")
```


**2. Scatterplot of GDP per capita vs. life expectancy by country in 1960**
```{r echo=FALSE}
df %>% filter(time==2019 & region!="") %>%
  ggplot(aes(x=GDP_PC, y=lifeExpectancy,color=region, size=totalPopulation)) + geom_point() + 
  xlab("GDP per capita") + ylab("Life Expectancy") + scale_size(guide="none") +
  guides(color=guide_legend(title="Continent")) +
  ggtitle("GDP per capita vs. life expectancy by country in 2019")
```



**3. Compare plots**


**4. Average life expectancy for each continent in 1960 and 2019**
```{r echo=FALSE}
df %>% filter(time==c(1960, 2016) & region!="") %>% 
  group_by(region, time) %>% 
  summarise(lifeExpectancy_avg = mean(lifeExpectancy, na.rm=TRUE)) %>%
  arrange(time)
```

**5. Compute average LE growht frmo 1960-2019 across continents**
```{r echo=FALSE}
df %>% group_by(region) %>% filter(region!="") %>%
  mutate(lifeExpectancy_prev = lag(lifeExpectancy), #Create a lag where we shift LE down by 1 year
         lifeExpectancy_growth = lifeExpectancy - lifeExpectancy_prev) %>% #Calculate the growth by subtracting LE - LE_prev
  summarise(lifeExpectancy_growth_avg = mean(lifeExpectancy_growth, na.rm=TRUE)) #take the average of LE_growth 
```





**6. Histogram of GDP per capita for years 1960 and 2019**
```{r echo=FALSE}
df %>% filter(time==c(1960, 2019)) %>%
  ggplot(aes(x=GDP_PC, fill=factor(time), color=factor(time))) + geom_histogram(alpha=0.4,colour = "black",
                 lwd = 0.75,
                 linetype = 1,) + 
  guides(fill=guide_legend(title="Year")) + xlab("GDP per capita")
```


**6. Histogram of GDP per capita for years 1960 and 2019**
```{r echo=FALSE}
df %>% filter(time==c(1960, 2019)) %>%
  ggplot(aes(x=GDP_PC, fill=factor(time), color=factor(time))) + geom_histogram(alpha=0.4,colour = "black",
                 lwd = 0.75,
                 linetype = 1,) + 
  guides(fill=guide_legend(title="Year")) + xlab("GDP per capita")
```



**7. Ranking of US in terms of life expectancy in 1960 and 2019**
```{r echo=FALSE}
df %>% filter(time==1960 & name!="") %>%
  mutate(rank=rank(desc(lifeExpectancy))) %>% filter(name == "United States of America") %>% select(name, rank)

df %>% filter(time==2019 & name!="") %>%
  mutate(rank=rank(desc(lifeExpectancy))) %>% filter(name == "United States of America") %>% select(name, rank)
```

**8. Relative rank**
```{r}
df %>% filter(time==1960 & name!="") %>%
  mutate(rank=rank(desc(lifeExpectancy))) %>% summarize(max_rank=max(rank))
```

